Calibration

### 2023-01-01 SingleSite_SingleDisturb.Rmd
Pushed results outline to own file, small updates to discussion and methods
Results: simulated data experiments

### Save outputs required to include figure in Results
save(data,CI.fN,time,an0.stats,an.stats,NT,Dbar,file="SSSD.RData")

>>>>

### Save outputs required to include figure in Results
save(data,CI.fN,time,an0.stats,an.stats,NT,Dbar,xlim,col,file="SSSD.RData")

## Conceptual Figure
```{r, echo=FALSE}
## params
set.seed(1066)
p.dist = 0.05
n0 = 1
nIC = 6
r = .1
K = 6
n.sd = 0.25
NE = 1000
obs=1
obs.sd = 0.75

## generate IC ensemble
ens = rnorm(NE,nIC,n.sd)

## Forecast step
LAI = ens + r*ens*(1-ens/K) + rnorm(NE,0,n.sd) 
disturbed = as.logical(rbinom(NE,1,p.dist))
LAI[disturbed] = abs(rnorm(sum(disturbed),n0,0.5)) ## disturbance

##Approximate forecast with Normal
xseq = seq(0,8,length=1000)
muf = mean(LAI)
Pf = var(LAI)

## analysis step: combine previous forecast with observed data
H <- matrix(1,ncol=1)
K <- Pf * t(H) %*% solve(H%*%Pf%*%t(H) + obs.sd^2)
mua <- muf + K*(obs - muf) 
Pa <- (1-K %*% H)*Pf

## EnKF
hist(LAI,probability = TRUE)#,ylim=c(0,dnorm(mua,mua,sqrt(Pa))))
lines(xseq,dnorm(xseq,muf,sqrt(Pf)),col="#b2df8a",lwd=3)
lines(xseq,dnorm(xseq,obs,obs.sd),col='red',lwd=3)
lines(xseq,dnorm(xseq,mua,sqrt(Pa)),col='orange',lwd=3)
legend("topleft",legend=c("forecast","data","assimilation"),lwd=3,col=c("#b2df8a","red","orange"))

## Forecast: Mixture of normals
mu1 = mean(LAI[!disturbed])
P1 = var(LAI[!disturbed])
mu2 = mean(LAI[disturbed])
P2 = var(LAI[disturbed])

## assimilation
priors <- list(p = p.dist,
                 mufN = mu1,
                 mufD = mu2,
                 pfN  = 1/P1,
                 pfD  = 1/P2,
                 Y = obs,
                 r = 1/obs.sd^2
                 )
mod <- jags.model(file=textConnection(AnalysisSSSD),
                   data=priors,n.adapt=1000,n.chains=3)
jdat <- coda.samples(mod,variable.names=c("n"),n.iter=10000)

hist(LAI,probability = TRUE,ylim=c(0,dnorm(mu1,mu1,sqrt(P1))))
lines(xseq,(1-p.dist)*dnorm(xseq,mu1,sqrt(P1)),col="#b2df8a",lwd=3)
lines(xseq,p.dist*dnorm(xseq,mu2,sqrt(P2)),col="brown",lwd=3)
lines(xseq,dnorm(xseq,obs,obs.sd),col='red',lwd=3)
lines(density(as.matrix(jdat)),col="blue",lwd=3)
legend("top",legend=c("undisturbed","disturbed","data","assimilation"),lwd=3,col=c("#b2df8a","brown","red","blue"))

#legend("top",legend=c("undisturbed","disturbed"),lwd=3,col=c("#b2df8a","brown"))

### 2023-01-01 methods.Rmd
Results: simulated data experiments

While this fusion of forecast and new data is an incredibly powerful concept, as alluded to in the _Introduction_ there are times that the Kalman Normal-Normal "nudge" produces results that don't always make sense. Consider as a simple example a simulated data experiment where LAI growth follows a simple stochastic logistic model with growth rate $r$=0.1, equilibrium $K$=6, process error standard deviation 0.25, and observation error standard deviation 0.75. Under normal conditions, EnKF would have no trouble nudging a LAI forecast back on track every time a new observation is made, but what if a discrete disturbance event occurs knocking LAI from steady-state down to $1.0 \pm 0.5$? In this case the compromise between the forecast ($LAI=6$) and observations ($LAI=1$) produces a compromise ($LAI \approx 5$ that is a poor representation of the state of the system (Figure \@ref(fig:enkf)). Furthermore, as the true LAI slowly recovers upward, the EnKF keeps gradually nudging LAI downward until the two converge. Not only is the assimilated state wrong, but so is our estimate of the process -- a discrete disturbance and rapid recovery is behaving like gradual degradation, which will lead to incorrect inferences about related processes (litter, soil C, NPP, NEE, ET, albedo, etc.). Not also that including a disturbance probability in the process model does not solve this problem. Assuming the disturbance probability is low (e.g. a background rates of <1%), at every point in time the forecast would include a small number of disturbed ensemble members, but the overwhelming majority would be undisturbed. Therefore $\mu_{f,t}$ will be relatively unchanged and $P_{f,t}$ would see a small increase in variance. The fundamental problem is that even though the true forecast is now bimodal (a large peak at the undisturbed state and a much smaller peak at the disturbed state), the EnKF is still representing this as a unimodal Normal. Indeed, if the you increased the disturbance probability substantially (e.g. 50%), the EnKF would treat this as a unimodal forecast half way between the disturbed and undisturbed states, a place where no ensemble members are present.

@@

@@

>>>>

While this fusion of forecast and new data is an incredibly powerful concept, as alluded to in the _Introduction_ there are times that the Kalman Normal-Normal "nudge" produces results that don't always make sense. Consider as a simple example a simulated data experiment where LAI growth follows a simple stochastic logistic model with growth rate $r$=0.1, equilibrium $K$=6, process error standard deviation 0.25, and observation error standard deviation 0.75. Under normal conditions, EnKF would have no trouble nudging a LAI forecast back on track every time a new observation is made, but what if a discrete disturbance event occurs knocking LAI from steady-state down to $1.0 \pm 0.5$? In this case the compromise between the forecast ($LAI=6$) and observations ($LAI=1$) produces a compromise ($LAI \approx 5$ that is a poor representation of the state of the system (Figure \@ref(fig:enkf)). Furthermore, as the true LAI slowly recovers upward, the EnKF keeps gradually nudging LAI downward until the two converge. Not only is the assimilated state wrong, but so is our estimate of the process -- a discrete disturbance and rapid recovery is behaving like gradual degradation, which will lead to incorrect inferences about related processes (litter, soil C, NPP, NEE, ET, albedo, etc.). Not also that including a disturbance probability in the process model does not solve this problem. Assuming the disturbance probability is low (e.g. a background rates of <1%), at every point in time the forecast would include a small number of disturbed ensemble members, but the overwhelming majority would be undisturbed. Therefore $\mu_{f,t}$ will be relatively unchanged and $P_{f,t}$ would see a small increase in variance. The fundamental problem is that even though the true forecast is now bimodal (a large peak at the undisturbed state and a much smaller peak at the disturbed state), the EnKF is still representing this as a unimodal Normal. Indeed, if the you increased the disturbance probability substantially (e.g. 50%), the EnKF would treat this as a unimodal forecast half way between the disturbed and undisturbed states, a place where no ensemble members are present (Figure \@ref(fig:concept), top).

@@

```{r concept, echo=FALSE, fig.cap="Top: In a conventional EnKF, a multimodal forecast (grey histogram) is approximated by a unimodal Normal (green), such that when a low probability disturbance is actually observed (red) the posterior (orange) does not capture this event. Bottom: In the multinomial filter the same multimodal forecast is represented by a mixture of normals (green and dark red) each weighted by their prior (forecast) probabulities. The posterior (blue), updates the mean and variance of each mode as well as their relative probabilities, in this case capturing the high probability that a disturbance occurred."}
## params
set.seed(1066)
p.dist = 0.05
n0 = 1
nIC = 6
r = .1
K = 6
n.sd = 0.25
NE = 1000
obs=1
obs.sd = 0.75

## generate IC ensemble
ens = rnorm(NE,nIC,n.sd)

## Forecast step
LAI = ens + r*ens*(1-ens/K) + rnorm(NE,0,n.sd) 
disturbed = as.logical(rbinom(NE,1,p.dist))
LAI[disturbed] = abs(rnorm(sum(disturbed),n0,0.5)) ## disturbance

##Approximate forecast with Normal
xseq = seq(0,8,length=1000)
muf = mean(LAI)
Pf = var(LAI)

## analysis step: combine previous forecast with observed data
H <- matrix(1,ncol=1)
K <- Pf * t(H) %*% solve(H%*%Pf%*%t(H) + obs.sd^2)
mua <- muf + K*(obs - muf) 
Pa <- (1-K %*% H)*Pf

## EnKF
hist(LAI, probability = TRUE, main = "Conventional EnKF")#,ylim=c(0,dnorm(mua,mua,sqrt(Pa))))
lines(xseq,dnorm(xseq,muf,sqrt(Pf)),col="#b2df8a",lwd=3)
lines(xseq,dnorm(xseq,obs,obs.sd),col='red',lwd=3)
lines(xseq,dnorm(xseq,mua,sqrt(Pa)),col='orange',lwd=3)
legend("topleft",legend=c("forecast","data","assimilation"),lwd=3,col=c("#b2df8a","red","orange"))

## Forecast: Mixture of normals
mu1 = mean(LAI[!disturbed])
P1 = var(LAI[!disturbed])
mu2 = mean(LAI[disturbed])
P2 = var(LAI[disturbed])

## assimilation
priors <- list(p = p.dist,
                 mufN = mu1,
                 mufD = mu2,
                 pfN  = 1/P1,
                 pfD  = 1/P2,
                 Y = obs,
                 r = 1/obs.sd^2
                 )
mod <- jags.model(file=textConnection(AnalysisSSSD),
                   data=priors,n.adapt=1000,n.chains=3)
jdat <- coda.samples(mod,variable.names=c("n"),n.iter=10000)

hist(LAI,probability = TRUE,ylim=c(0,dnorm(mu1,mu1,sqrt(P1))), main="Multinomial Filter")
lines(xseq,(1-p.dist)*dnorm(xseq,mu1,sqrt(P1)),col="#b2df8a",lwd=3)
lines(xseq,p.dist*dnorm(xseq,mu2,sqrt(P2)),col="brown",lwd=3)
lines(xseq,dnorm(xseq,obs,obs.sd),col='red',lwd=3)
lines(density(as.matrix(jdat)),col="blue",lwd=3)
legend("top",legend=c("undisturbed","disturbed","data","assimilation"),lwd=3,col=c("#b2df8a","brown","red","blue"))
```

@@

Divergence

As a metric to assess the prevalence of filter divergence, we calculated the difference between the last observed data point and the posterior mean of the corresponding Analysis, and then divided this by half the width of the Analysis 95% confidence interval. If the confidence interval is symmetric, then any Analysis mean that falls in the confidence interval will score between -1 and 1, while values outside this range are suggestive of filter divergence.

### 2023-01-01 results.Rmd
Results: simulated data experiments

1. Simulated data experiments (known disturbance)
  + Single Site, Single Disturbance

Figures showing the current problem, solution, and that we've got a genuine multimodal distribution and P(dist)

  + SSSD Multi Pool

```{r, echo=FALSE,fig.cap="Experiment 2: Assimilation using process-based VSEM captures multi-pool disturbance to leaf and wood biomass"}
load("exp2.Rdata") # Generated by SingleSite_SingleDisturb_MultiPool.Rmd
#Figures showing all 3 pools cov, C accounting / attribution 
pool.names=c("Cleaf","Csoil","Cwood")
for(i in 1:3){

@@

```{r,echo=FALSE, fig.asp=1,fig.width=3, fig.cap="Experiment 2: Forecast and Analysis ensembles at the disturbance (t=10). Arrows demonstrate the ensemble adjustment of a sample of 10 ensemble members."}
load("exp2.Rdata")

@@

  + SS Multi-Disturbance w/ PFT switching

Figure showing PFT switching vs not, ability to detect disturbance type

  + Simple Multi-site: spatial cov

Figure showing "dragging down" of neighbors in conventional and breaking of COV in Multinomial

```{r, echo=FALSE,fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance"}

## figure with 3 rows, 2 col showing timeseries in rt and bivariate in left

@@

## ensemble adjustment
ne = nrow(Forecast[[10]]$Nf)
sel.a= seq(1,ne,length=10)#sample.int(ne,10)
arrows(Forecast[[10]]$Nf[sel.a,1],Forecast[[10]]$Nf[sel.a,3],
       Analysis[[10]]$Na[sel.a,1],Analysis[[10]]$Na[sel.a,3],length = 0.1,lwd = 0.5)   
legend("topleft",
       legend=c("Forecast","Analysis","Data"),
       pch=c(1,3,18),col=c("purple","#b2df8a",1),
       cex = 1.5)

@@

## ensemble adjustment
ne = nrow(Forecast[[10]]$Nf)
sel.a= seq(1,ne,length=10)#sample.int(ne,10)
arrows(Forecast[[10]]$Nf[sel.a,1],Forecast[[10]]$Nf[sel.a,3],
       Analysis[[10]]$Na[sel.a,1],Analysis[[10]]$Na[sel.a,3],length = 0.1,lwd = 0.5)   
legend("topleft",
       legend=c("Forecast","Analysis","Data"),
       pch=c(1,3,18),col=c("purple","#b2df8a",1),
       cex = 1.5)

@@

@@

2. Real-world examples
  + Calibrate state-and-transition prior
    * data: (samples from) maps of disturbance type by year
    * model: Multinomial Dirlichet

Figure showing map and estimated priors

  + Applications
    + Logging
    + Fire
    + Land use transition (forest -> non?)

* RESULTS
  * needed to reduce obs and process error to work
  * abs magnitude of disturbance is best predictor of assim prob.
    * given magnitude, dist type n.s.

>>>>

Overall, the results from the simulated data experiments (Table \@ref(tab:pseudoTable)) demonstrate that the proposed multinomial ensemble filter is able to successfully detect disturbances under a range of different conditions.

Experiment 1, which aimed to capture a single disturbance at a single site for a single pool, was already depicted in Figure \@ref(fig:enkf) and Figure \@ref(fig:concept) when discussing the challenges facing traditional Kalman Filter variants. Specifically, Figure \@ref(fig:enkf) shows that the multinomial filter (blue) is able to capture the discrete nature of the disturbance at time 10, with the posterior probability of disturbance jumping from near zero before this event to 0.977 for the disturbance, and then dropping back down to near zero in subsequent years. Figure \@ref(fig:concept) bottom illustrates what is occurring in this disturbance year. As in the prior 9 years the forecast itself is multimodal, with most of the ensemble members predicting no disturbance and a small number predicting a disturbance event. The multinomial filter assigns a prior mean and variance to each mode based on the ensemble summary statistics, which is the same as a traditional EnKF except for the stratification by disturbance type, and then assigns each state a prior probability based on the ensemble frequency of being in each state (disturbed = dark red, undisturbed = green). When the observed data (red) suggests a large drop in LAI, the posterior (blue) not only updates the ensemble mean and variance of each mode, but also the probability of being in each mode. This posterior is still multimodal, but in this specific instance the weight assigned the undisturbed mode (0.023) is too low to be visually notable. Noteworthy is that the LAI in years 12 and 15 is actually lower than that in the disturbance year but the multinomial, yet the model still assigns this as being undisturbed. This occurs because in these years the observed low LAI is compatible with the observation error around the undisturbed forecast, which has a higher prior probability than the odds of subsequent disturbance.

Experiment 2 expands the assimilation to a multivariate forecast, in this case including joint predictions for leaf, wood, and soil carbon, and with a disturbance to both the leaf and wood pools (Figure \@ref(fig:sssd)). Figure \@ref(fig:sssdBi) shows the multivariate nature of the detection of the disturbance event at time 10, with the bimodal forecast ensemble shown in purple (undisturbed in the top right, disturbed in the bottom left), an observation in black that is consistent with the disturbed state, and the posterior ensemble in green. Arrows depict the adjustment of a random subset of individual ensemble members from the forecast prior to the analysis posterior.

```{r sssd, echo=FALSE,fig.cap="Experiment 2: Assimilation using process-based VSEM captures multi-pool disturbance to leaf and wood biomass"}
load("exp2.Rdata") # Generated by SingleSite_SingleDisturb_MultiPool.Rmd
#Figures showing all 3 pools cov, C accounting / attribution 
pool.names=c("Cleaf","Csoil","Cwood")
for(i in 1:3){

@@

```{r sssdBi,echo=FALSE, fig.asp=1,fig.width=3, fig.cap="Experiment 2: Forecast and Analysis ensembles at the disturbance (t=10). Arrows demonstrate the ensemble adjustment of a sample of 10 ensemble members."}
load("exp2.Rdata")

@@

Experiment 3 (Figure \@ref(fig:multiPseudo) top row) illustrates the ability of the multinomial model to select between more than one disturbance. In the bivariate plot (top right panel) we can see that the model can successfully distinguishing between two types of disturbance, one that involves the loss of both wood and leaf biomass (e.g. fire, logging) and the other that just removes leaf biomass (e.g. defoliating forest pest). Experiment 4 (Figure \@ref(fig:multiPseudo) middle row) is similar to 3 but extends the woody biomass removing disturbance to one that causes a switch in PFT from a forest to a low-stature vegetation type (e.g. grassland, shrubland, agriculture), demonstrating that the model can handle land use and land cover changes. Finally, experiment 5 (Figure \@ref(fig:multiPseudo) bottom row) illustrates the extension of the assimilation to multiple sites, one that experiences and disturbance and one that does not. In the years prior to the disturbance the multivariate nature of the assimilation allows us to borrow strength across sites, with observations at one site reducting uncertainty not just at that site but at the other site as well based on the cross-site forecast covariance. That said,this example illustrates the ability of the multinomial model to avoid the impact of spurious correlations across sites during disturbance events. The presence of the disturbance at site 1 did not cause a reduction in biomass in site 2 during the disturbance, nor did it increase the productivity in site 2 in the years post-disturbance despite the large increase observed at site 1.

```{r multiPseudo, echo=FALSE,fig.cap="Simulated experiments 3-5: Left hand column shows timeseries for the wood pool for all three experiments; Right hand shows bivariate ensemble adjustment at time of disturbance",fig.height=4,fig.width=3}

## figure with 3 rows, 2 col showing timeseries in rt and bivariate in left
#par(mfrow=c(3,2))
m = matrix(c(1,1,2,3,3,4,5,5,6),nrow=3,byrow = TRUE)
layout(m)

@@

## ensemble adjustment
ne = nrow(Forecast[[10]]$Nf)
sel.a= seq(1,ne,length=10)#sample.int(ne,10)
arrows(Forecast[[10]]$Nf[sel.a,1],Forecast[[10]]$Nf[sel.a,3],
       Analysis[[10]]$Na[sel.a,1],Analysis[[10]]$Na[sel.a,3],length = 0.1,lwd = 0.5)   
legend("bottomright",
       legend=c("Forecast","Analysis","Data"),
       pch=c(1,3,18),col=c("purple","#b2df8a",1),
       cex = 1.25)

@@

plot(Analysis[[10]]$Na[,1],Analysis[[10]]$Na[,3],pch=Analysis[[10]]$Ca,col=ecoforecastR::col.alpha("#b2df8a",alpha=1),
     ylim=range(Forecast[[10]]$Nf[,3]),
     xlim=range(Forecast[[10]]$Nf[,1]),
     xlab=pool.names[1],ylab=pool.names[3],
     cex.lab=1.25)
points(Forecast[[10]]$Nf[,1],Forecast[[10]]$Nf[,3],cex=1,col="purple",pch=18)
points(No[10,1],No[10,3],cex=3,col=1)
lines(No[10,1]+c(-1,1)*sqrt(R[1,1]),rep(No[10,3],2))
lines(rep(No[10,1],2),No[10,3]+c(-1,1)*sqrt(R[3,3]))

## ensemble adjustment
ne = nrow(Forecast[[10]]$Nf)
sel.a= seq(1,ne,length=10)#sample.int(ne,10)
arrows(Forecast[[10]]$Nf[sel.a,1],Forecast[[10]]$Nf[sel.a,3],
       Analysis[[10]]$Na[sel.a,1],Analysis[[10]]$Na[sel.a,3],length = 0.1,lwd = 0.5)   
legend("bottomright",
       legend=c("Forecast","Analysis","Data"),
       pch=c(1,3,18),col=c("purple","#b2df8a",1),
       cex = 1.25)

@@

## VSEM Calibration

@@

```{r ORorig, echo=FALSE}
## summary figure showing didn't work great at default error rate

### Row 1: Example time-series ###
site = 188
exp.name = c("P1Q1R1","P1Q1R4")
exp.pretty.name = c("Original","Low Obs Error")
pool.names=c("Cleaf","Csoil","AGB")
i = 3 ## Stem
for(ex in seq_along(exp.name)){
  load(paste0("OR_TS_",site,"_",exp.name[ex],".RData"))
  plot(Yi[,i],
       ylim = range(rbind(CI.Na,Yi[,i],CI.D),na.rm=TRUE),
       type='n',
       ylab=pool.names[i],xlab="time",
       main=exp.pretty.name[ex])
  ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[5,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  lines(CI.Na[3,],col="#b2df8a",lwd=2)
  points(Yi[,i])
  if(ex == 1) {
    legend("bottomleft",legend=c("Assimilation","Data"),
       lwd=5,col=c("#b2df8a","black"),cex=0.7)
  }
}

### Row 2: Divergence ###
load("stats.RData")
load("elist.RData")
disturb.names = c("Intact","Other","Cut","Fire","Pest")
disturb.col = c("#ebebeb",'#b04c00','#1f78b4','#faabaa','#b2df8a')
disturb.col.al <- tapply(disturb.col,1:5,FUN=ecoforecastR::col.alpha,0.7); disturb.col.al[1] = disturb.col[1]
for(i in 1:2){
  ## divergence
  div = (elist[[i]]$divergence[,1] - obs.tlo)/
    (apply(elist[[i]]$divergence[,c(2,6)],1,diff)*0.5)
  div.reg    = lm(div~obs.tlo)

  dh = density(na.omit(div))
#  abline(v=c(-1,1),lty=2)
  plot(obs.tlo,div,col=disturb.col.al[obs.class],
       xlab="final AGB",ylab="Divergence",
       pch=16)
  abline(h=c(-1,1),lty=2)
  abline(div.reg,col="purple",lwd=3)
  abline(v=0,lty=2)
  lines(dh$y*10,dh$x,lwd=3)
  if(i == 1) legend("topright",legend=disturb.names,
                  col=disturb.col,pch=16,cex=0.7,
                  text.col=c(1,disturb.col[2:5]))
}

### Row 3: Disturbance magnitude vs P(detection) ###
pred.seq=list()
distabs = obs.pre-obs.dist
pred.seq$distabs = seq(min(distabs),max(distabs),length=500)
for(i in 1:2){
  distrate = distrate = elist[[i]]$distrate
  dist.qb.abs = glm(distrate ~ distabs,
                    na.action=na.omit,
                    family=quasibinomial("logit")
  )
  plot(distabs,distrate,col=disturb.col.al[obs.class],
       xlab="AGB disturbance magnitude",ylab="P(detect)",
       pch=16)
  lines(pred.seq$distabs,predict(dist.qb.abs,newdata=pred.seq,type="response"),col="purple")
  if(i == 1) legend("topleft",legend=disturb.names,
                  col=disturb.col,pch=16,cex=0.7,
                  text.col=c(1,disturb.col[2:5]))
}
```

* RESULTS
  * limited success under default parameters
    * high frequency of filter divergence [summary fig, example divergence fig]
    * divergence coverage original `r stats[[1]]$div.cov`
    * divergence coverage low error `r stats[[2]]$div.cov`
    * the smaller the observed biomass at the end of the timeseries, the larger the divergence [include scatterplot?]
    * low rate of detected disturbance [calc stat?, hist? cdf?]
    * what disturbances were detected tended to be associated with high pre-disturbance biomass and a large reduction in biomass (both absolute and proportional). Most commonly seen in dist class "red"
  * needed to reduce obs and process error to work
  * example "successful" scenario: P1Q1R4
    * reduced observation error
    * abs magnitude of disturbance is best predictor of assim prob. [fig]
      * given magnitude, dist type n.s.  
  * comparative analysis
